---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r echo = F}

knitr::opts_chunk$set(echo = T)

```
# Loading and Preprocessing the Data

I download the activity dataset from the link given by the assignment. I then unzip the file, and read it in. I also load required packages that will be used.

```{r}
library(ggplot2, quietly = T)
library(data.table, quietly = T)

download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
              destfile = "activity.zip")

unzip("activity.zip")
```

The activity file was downloaded on `r Sys.time()`. I will use the data.table package to do the required calculations. 

```{r}
dat = read.csv("activity.csv")

dat = data.table(dat)
setkey(dat,date)
```

The dataset has the following format
```{r}
summary(dat)
```

# Performing Calculations

## What is the Mean Total Number of Steps Taken Per Day?

Lets calculate the total number of steps taken per day
```{r}
totalstepsbyday = dat[,list("total_steps" = sum(steps,na.rm = T)), by = date]
```

Let's look at a histogram of the total number of steps per day.
```{r}
qplot(total_steps, data = totalstepsbyday, geom = "histogram",
      main = "Histogram of Total Number of Steps per Day\n")
```

Let's calculate the mean and the median of total number of steps per day.
```{r}
totalstepsbyday[,list("median number of steps per day" = median(total_steps,na.rm = T),
          "mean number of steps per day" = mean(total_steps,na.rm = T))]
```

## What is the Average Daily Activity Pattern

Lets explore the average number of steps taken, averaged accross all days.
```{r}
avgstepsperinterval = dat[,list("avg_steps_interval" = mean(steps, na.rm = T)), by = interval]

qplot(interval, avg_steps_interval, data = avgstepsperinterval, main = "Average Steps By Interval\n", xlab = "Interval", ylab = "Average Number of Steps", geom = c("line"))
```

The `r avgstepsperinterval[which(avg_steps_interval == max(avg_steps_interval)),interval]` interval has the has on average the maximum number of steps
```{r}
avgstepsperinterval[which(avg_steps_interval == max(avg_steps_interval)),]

```

## Imputing Missing Values

The total number of missing values by each field is
```{r}
colSums(is.na(dat))
```
This means there are `r sum(colSums(is.na(dat)))` rows out of `r nrow(dat)` that have missing data.

Will fill in those missing values for each data point by using the average number of steps for that interval.
```{r}
avgstepsperinterval = data.table(avgstepsperinterval)
setkey(avgstepsperinterval,interval)

imputeddat = copy(dat)
# missing values in steps
missingStepsData = is.na(imputeddat$steps)
# convert steps to double so that we can assign the mean to missing values
imputeddat[, steps := as.double(steps)]
imputeddat[missingStepsData,steps := (avgstepsperinterval[.(interval),avg_steps_interval])]

# A spot check for some values to ensure the assignment was correct

avgstepsperinterval[.(0)]
imputeddat[missingStepsData & interval == 0]
# interval 0 looks correct

avgstepsperinterval[.(35)]
imputeddat[missingStepsData & interval == 35]
# interval 35 looks correct

avgstepsperinterval[.(20)]
imputeddat[missingStepsData & interval == 20]
# interval 20 looks correct
```

Lets calculate the total number of steps taken per day after imputing missing values.
```{r}
imputedtotalstepsbyday = imputeddat[,list("total_steps" = sum(steps,na.rm = T)), by = date]
```

Let's look at a histogram of the total number of steps per day aftering imputing missing values.
```{r}
qplot(total_steps, data = imputedtotalstepsbyday, geom = "histogram",
      main = "Histogram of Total Number of Steps per Day\n")
```

Let's calculate the mean and the median of total number of steps per day after imputing missing values.
```{r}
imputedtotalstepsbyday[,list("median number of steps per day" = median(total_steps,na.rm = T),
          "mean number of steps per day" = mean(total_steps,na.rm = T))]
```

## Are there Differences In Activity Patterns Between Weekdays and Weekends?

We will explore differences in activity patterns between weekdays and weekends. 

Start by creating a factor nothing whether a day is a weekday or weekend, and then plot the average number of steps taken per interval, on weekdays and weekends.

```{r}
imputeddat[,dayofweek := weekdays(as.Date(date))]
imputeddat[, type := factor(dayofweek == "Saturday" | dayofweek == "Sunday", labels = c("Weekday","Weekend"))]

imputedavgnumstepsbydaytype = imputeddat[,.("avg_num_steps" = mean(steps, na.rm = T)), keyby = .(type, interval)]

qplot(x = interval, y = avg_num_steps, data = imputedavgnumstepsbydaytype, facets = type ~ ., geom = "line")
```